\bf The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-09 paper/forestecology.tex
About to start slash & burn 712 words

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evalulate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We highlight the
importance of spatial cross validation when interpreting model
results.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, interspecific competition, neigbhorhood competition,
tree growth, R, ForestGEO, spatial cross validation
\vfill

@@

Repeat-censused forest plots offer excellent opportunities to test
neighborhood models of the effect of competition on the growth of trees
(\citet{canham_neighborhood_2004}). Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters (\citet{uriarte_spatially_2004}); measure species-specific
competition coefficients (\citet{das_effect_2012}
\citet{tatsumi_estimating_2016}); test competing models to see what
structures competitive interactions, e.g.~traits or phylogeny
(\citet{allen_permutation_2020}; \citet{uriarte_trait_2010}); and inform
selective logging practices (\citet{canham_neighborhood_2006}). Although
these are well-described methods, no methods are currently available for
their easy application. Here we address this in an R package. We largely
follow the methods presented in \citet{allen_permutation_2020}. The
package is written to model stem radial growth between two censuses
based on neighborhood competition.

\citet{allen_permutation_2020} considers the following model: Let
\(i = 1, \ldots, n_j\) index all \(n_j\) trees of ``focal'' species
group \(j\); let \(j = 1, \ldots, J\) index all \(J\) focal species
groups; and let \(k = 1, \ldots, K\) index all \(K\) ``competitor''
species groups. We model the average annual growth in diameter
\(y_{ij}\) (in centimeters per year) of the \(i^{th}\) tree of focal
species group \(j\) as a linear model \(f\) of the covariates
\(\vec{x}_{ij}\)

@@

where \(\beta_{0,j}\) is the diameter-independent growth rate for group
\(j\); \(\text{dbh}_{ij}\) is the diameter at breast height (in
centimeters) of the focal tree at the earlier census;
\(\beta_{\text{dbh},j}\) is the amount of the growth rate changed
depending on diameter for group \(j\); \(\text{BA}_{ijk}\) is the sum of
the basal area of all trees of competitor species group \(k\);
\(\lambda_{jk}\) is the change in growth for individuals of group \(j\)
from nearby competitors of group \(k\); and \(\epsilon_{ij}\) is a
random error term distributed \(\text{Normal}(0, \sigma^2)\). They
estimate all parameters via Bayesian linear regression while exploiting
Normal/Inverse Gamma conjugacy to derive closed-form solutions to all
posterior distributions via linear algebra\footnote{See S1 Appendix of
  \citet{allen_permutation_2020}, available at
  \url{https://doi.org/10.1371/journal.pone.0229930.s004}}. These
closed-form solutions for the posterior distributions are in contrast to
approximations of all posteriors via computationally expensive Markov
Chain Monte Carlo algorithms.

@@

We introduce the \texttt{forestecology} R package providing methods and
data for forest ecology model fitting and assessment, available on CRAN
(\url{https://cran.r-project.org/web/packages/forestecology/index.html})
with the corresponding source code available on GitHub
(\url{https://github.com/rudeboybert/forestecology}). The package
implements all aspects of the model in Equation \ref{eq:model}: model
fitting and generating fitted/predicted values, evaluating the effect of
competitor species identity using permutation tests, and evaluating
model performance using spatial cross-validation.

@@

Currently the package only implements the Bayesian linear regression
model of tree growth based on neighborhood competition detailed in
Equation \ref{eq:model}. As we demonstrate in Section
\ref{model-fit-predict} however, the fitting of this model is
self-contained in a single function \texttt{comp\_bayes\_lm()}. This
function returns an object of S3 class type \texttt{comp\_bayes\_lm}
with generic methods implemented for \texttt{print()} to inspect the
output, \texttt{predict()} to generate fitted/predicted values, and
\texttt{ggplot2::autoplot()} to visualize all results. Therefore the
package can be modularly extended to fit other models as long as they
are coded into a function similar type as \texttt{comp\_bayes\_lm()} as
has equivalent generic methods implemented.

@@

We demonstrate the \texttt{forestecology} package's functionality on
data from the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot, located at the Smithsonian's National Zoo and
Conservation Biology Institute in Front Royal, VA, USA
\citep{bourg_initial_2013}. The 25.6 ha (640 x 400 m) plot is located at
the intersection of three of the major physiographic provinces of the
eastern US---the Blue Ridge, Ridge and Valley, and Piedmont
provinces---and is adjacent to the northern end of Shenandoah National
Park. The forest type is typical mature secondary eastern mixed
deciduous forest, with a canopy dominated by tulip poplar
(\emph{Liriodendron tulipifera}), oaks (\emph{Quercus} spp.), and
hickories (\emph{Carya} spp.), and an understory composed mainly of
spicebush (\emph{Lindera benzoin}), paw-paw (\emph{Asimina triloba}),
American hornbeam (\emph{Carpinus caroliniana}), and witch hazel
(\emph{Hamamelis virginiana}) \citep{bourg_initial_2013}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(lubridate)}
\KeywordTok{library}\NormalTok{(sf)}
\KeywordTok{library}\NormalTok{(patchwork)}
\KeywordTok{library}\NormalTok{(forestecology)}
\KeywordTok{library}\NormalTok{(blockCV)}
\end{Highlighting}
\end{Shaded}

\hypertarget{step-1-compute-the-growth-of-trees-based-on-census-data-compute-growth}{%
\subsection{Step 1: Compute the growth of trees based on census data
\{compute-growth\}}\label{step-1-compute-the-growth-of-trees-based-on-census-data-compute-growth}}

The first step in the our analysis sequence is to compute the growth of
trees using data from two censuses. The \texttt{compute\_growth()}
function computes average annual growth assuming census data that
roughly follows ForestGEO standards. Despite such standards, minor
variations will still exist between sites, thereby necessitating some
data wrangling and checking. For example, the SCBI site records all
diameters at breast height (DBH) in millimeters
\citep{bourg_initial_2013}, whereas the Michigan Big Woods site records
them in centimeters \citep{allen_michigan_2020}.

We load both 2008 and 2014 SCBI census data \texttt{.csv} files as they
existed on GitHub on November 20, 2020
\citep{gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020}. After
selecting only the relevant variables, we perform a few additional data
wrangling steps: convert the character variable with the date of
measurement to be of explicit type \texttt{date}, convert DBH to be in
centimeters\footnote{A rule of thumb to determine the units of DBH is to
  check if the smallest non-zero and non-missing measurement is 1 or 10.
  If the former, then centimeters. If the later, then millimeters. This
  is because ForestGEO protocols state that only trees with DBH greater
  or equal to 1cm should be included in censuses.} Furthermore, in order
to speed up computation for purposes of this example, we only consider a
9 ha subsection of the 25.6 ha of the SCBI site: \texttt{gx} from 0--300
instead of 0--400 and \texttt{gy} from 300--600 instead of 0--640.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{census_}\DecValTok{2013}\NormalTok{_scbi <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"scbi.stem2.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(stemID, sp, }\DataTypeTok{date =}\NormalTok{ ExactDate, gx, gy, dbh, codes, status) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{date =} \KeywordTok{mdy}\NormalTok{(date),}
    \DataTypeTok{dbh =} \KeywordTok{as.numeric}\NormalTok{(dbh)}\OperatorTok{/}\DecValTok{10}
\NormalTok{  ) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(gx }\OperatorTok{<}\StringTok{ }\DecValTok{300}\NormalTok{, }\KeywordTok{between}\NormalTok{(gy, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{))}

@@

These two data frames are then supplied as arguments to the
\texttt{compute\_growth()} function, along with the \texttt{id} argument
that specifies the variable that uniquely identifies each tree-stem.
Note furthermore that we discard all resprouts in the later census
(those with \texttt{code\ ==\ R}), since we are only interested in the
diameter growth of surviving, and not resprouted, stems.

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes variables \texttt{growth}, the average annual
growth in DBH (cm y\textsuperscript{-1}) for all stems that were alive
at both time points, and \texttt{geometry}, the \texttt{sf} package's
encoding of geolocations of type
\texttt{\textless{}POINT\textgreater{}}. In addition the species
variable \texttt{sp} is converted to a factor if it wasn't already by
\texttt{compute\_growth()}.\footnote{In our spatial cross-validation
  algorithm in Section \ref{spatial-cross-validation} issues can occur
  when rare species do not occur in the training set, but then are
  encountered in the test set. This risk is mitigated by representing
  \texttt{sp} as a factor variable, which has a complete list of all
  levels of the categorical variable.} Furthermore, the variables that
should remain unchanged between censuses appear only once, such as
location variables \texttt{gx} and \texttt{gy}; as well as
species-related variables. Variables that should change between censuses
are suffixed with \texttt{1} and \texttt{2} indicating the earlier and
later censuses, such as \texttt{dbh1/dbh2} and \texttt{codes1/codes2}.

Site data that does not align with this convention will need to be
transformed for use with the \texttt{compute\_growth()} function.
However, in the end, all that matters is that the growth of all stems is
saved in a data frame of class \texttt{sf} and, at a minimum, contains
the variable uniquely identifying each stem, \texttt{sp}, \texttt{dbh1},
\texttt{growth}, \texttt{geometry}.

@@

The next step in our analysis sequence is to add additional spatial
information to our main \texttt{growth\_scbi} data frame. The first
element of spatial information we add is a ``buffer region'' to the
periphery of the study region. Since some of our model's explanatory
variables are cumulative (such as competitor basal area), we must ensure
that all trees being modeled are not biased to have different neighbor
structures. This is of concern for trees at the boundary of study
regions, for which all neighbors will not be included in the censused
stems. In order to account for such edge effects, only trees that are
not part of this buffer region, i.e.~are part of the interior of the
study region, will have their growth modeled
\citep{waller_applied_2004}.

Our model of interspecific competition relies on a spatial definition of
who the competitor trees are for focal trees of interest: all trees
within a distance \texttt{comp\_dist} of a focal tree are considered its
competitors (assuming the same units as the \texttt{gx} and \texttt{gy}
location variables). In our case we set this value at 7.5m, a value
informed by other studies \citep[\citet{uriarte_spatially_2004},
\citet{canham_neighborhood_2006}]{canham_neighborhood_2004}. Using this
value along with a manually constructed \texttt{sf} object
representation of the study region's boundary via its vertices, we apply
the \texttt{add\_buffer\_variable()} to our \texttt{growth\_scbi} data
frame to add a \texttt{buffer} boolean variable: all trees who have
\texttt{buffer} set to \texttt{FALSE} will be our focal trees whose
growth will be modeled, whereas those with \texttt{TRUE} will only be
considered as competitor trees whose growth will not.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Define buffer region using competitive distance range}
\NormalTok{comp_dist <-}\StringTok{ }\FloatTok{7.5}

\NormalTok{study_region_scbi <-}\StringTok{ }\KeywordTok{tibble}\NormalTok{(}
  \DataTypeTok{x =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{300}\NormalTok{, }\DecValTok{300}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{),}
  \DataTypeTok{y =} \KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{, }\DecValTok{600}\NormalTok{, }\DecValTok{300}\NormalTok{)}
\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{sf_polygon}\NormalTok{()}

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate
out-of-sample model error. Conventional cross-validation algorithms
assign observations to folds by randomly resampling individual
observations. However, many of these algorithms assume that the
observations are independent of each other. In the case of forest census
data, observations exhibit spatial autocorrelation. We therefore
incorporate this spatial dependence into the cross-validation algorithm
with our spatial blocks of trees \citep[
\citet{pohjankukka_estimating_2017}]{roberts_cross-validation_2017} In
the example below, we first manually define four folds that partition
the study region as an \texttt{sf} object. We then use the output of the
\texttt{spatialBlock()} function from the \texttt{blockCV} package to
associate each tree in \texttt{growth\_scbi} to the correct fold (saved
in the \texttt{foldID} variable) \citep{valavi_blockcv_2019}. \footnote{In
  the Supporting Information we present an example where the folds
  themselves are also created using the \texttt{spatialBlock()} function
  given a specified \texttt{cv\_block\_size}.}

@@

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of
adding these two elements of information to the \texttt{growth\_scbi}
data frame. The location of each tree is marked with an integer
indicating which fold it belongs to, where the folds are marked with
solid lines. The color of each digit indicates whether the tree is part
of the buffer region (and thus will only be considered as a competitor
tree in our model) or is part of the interior of the study region (and
thus is a focal tree whose growth is of modeled interest).

@@

The next step in our analysis sequence is to identify all focal trees
and their corresponding competitor trees. More specifically, it
identifies all trees that are not part of the buffer region, have a
valid \texttt{growth} measurement, and have at least one neighbor within
7.5 of it. The \texttt{create\_focal\_vs\_comp()} functions performs
these tasks and returns a new data frame of type \texttt{sf}. On top of
the previous arguments \texttt{comp\_dist} defining the competition
neighborhood and \texttt{id} indicating which variable uniquely
identifies each tree-stem, this function also requires an \texttt{sf}
object representation of the spatial cross-validation blocks/folds. In
this example, the blocks were manually encoded in \texttt{blocks\_scbi}
by specifying it's vertices in Section
\ref{spatial-information}\footnote{We present an alternative method for
  defining spatial cross-validation blocks is using the
  \texttt{spatialBlock()} function from the \texttt{blockCV} package in
  the Supporting Information.}. We present the resulting data frame
below with the \texttt{foldID} variable omitted for compactness of
presentation.

@@

The resulting data frame \texttt{focal\_vs\_comp\_scbi} has 6296 rows,
representing the subset of the 7954 trees in \texttt{growth\_scbi} that
will be considered as focal trees. Two new variables \texttt{focal\_ID}
and \texttt{focal\_sp} relate to tree-stem identification and species
information. Most notably however is a new variable \texttt{comp} which
contains information on all competitor trees for a given focal tree,
saved in \texttt{tidyr} package list-column format
\citep{tidyr_package}. For example, we drill-down on the tree with
\texttt{focal\_ID} 4, which has 20 competitor trees each described by 4
variables as indicated by the fact that \texttt{comp} is a
\texttt{tibble\ {[}20\ ×\ 4{]}}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(focal_ID }\OperatorTok{==}\StringTok{ }\DecValTok{4}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(focal_ID, dbh, comp)}
\CommentTok{## # A tibble: 1 x 3}
\CommentTok{##   focal_ID   dbh comp             }
\CommentTok{##      <dbl> <dbl> <list>           }
\CommentTok{## 1        4  13.6 <tibble [20 x 4]>}
\end{Highlighting}
\end{Shaded}

The spatial distribution of these trees is visualized in Figure
\ref{fig:scbi-focal-vs-comp-map}: the dashed circle extends 7.5 m away
from the focal tree while all 20 competitor trees are within this
circle.

@@

Using the \texttt{unnest()} function from the \texttt{tidyr} package, we
can flatten list-column into regular columns. We observe that for the
same focal tree, we have information on all 20 competitor trees whose
\texttt{dist} distance to the focal tree is \(\leq\) 7.5: their unique
tree-stem ID number, their species, and their basal area (in m\(^2\))
calculated as \(\frac{\pi \times (\text{DBH/2})^2}{10000}\) where
\(DBH\) is the value from the earlier of the two censuses in cm. Saving
our focal versus competitor information in list-column minimizes
redundancy since we do not repeat information on the focal tree 20
times.

@@

Now that we've identified all focal and corresponding competitor trees
and saved this information in a data frame of type
\texttt{focal\_vs\_comp}, the final step in our analysis sequence is to
fit a model for the growth of all focal trees. Currently the
\texttt{forestecology} package can only fit the competition Bayesian
linear regression model outlined in Section \ref{competition-model}
using the \texttt{comp\_bayes\_lm()} function. However, any model
implemented in a function that similarly takes an input data frame of
type \texttt{focal\_vs\_comp} as an argument can also be used. For our
specific competition Bayesian linear regression model, we also specify
prior distributions on all parameters of interest (here chosen to be the
defaults as specified in \texttt{?comp\_bayes\_lm}).

@@

The returned \texttt{comp\_bayes\_lm\_scbi} output is an object of S3
class type \texttt{comp\_bayes\_lm} which contains the posterior values
of all parameters in our competition Bayesian linear regression. This
class of object includes generic methods implemented for
\texttt{print()}, \texttt{predict()}, and \texttt{ggplot2::autoplot()}.
First the generic for \texttt{print()} displays the names of all prior
\& posterior parameters along with the model formula:

>>>>

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evalulate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We highlight the
importance of spatial cross-validation when interpreting model
results.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, interspecific competition, neigbhorhood competition,
tree growth, R, ForestGEO, spatial cross-validation
\vfill

@@

Repeat-censused forest plots offer excellent opportunities to test
neighborhood models of the effect of competition on the growth of trees
(\citet{canham_neighborhood_2004}). Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters (\citet{uriarte_spatially_2004}); measure species-specific
competition coefficients (\citet{das_effect_2012}
\citet{tatsumi_estimating_2016}); test competing models to see what
structures competitive interactions, e.g.~traits or phylogeny
(\citet{allen_permutation_2020}; \citet{uriarte_trait_2010}); and inform
selective logging practices (\citet{canham_neighborhood_2006}). Although
these are well-described methods, few methods are currently available
for easy application. Here we address this in an R package. We largely
follow the methods presented in \citet{allen_permutation_2020}. The
package is written to model stem radial growth between two censuses
based on neighborhood competition.

\citet{allen_permutation_2020} considers the following model: Let
\(i = 1, \ldots, n_j\) index all \(n_j\) trees of ``focal'' species
group \(j\); let \(j = 1, \ldots, J\) index all \(J\) focal species
groups; and let \(k = 1, \ldots, K\) index all \(K\) ``competitor''
species groups. We model the average annual growth in diameter at breast
height (DBH) \(y_{ij}\) (in centimeters per year) of the \(i^{th}\) tree
of focal species group \(j\) as a linear model \(f\) of the covariates
\(\vec{x}_{ij}\)

@@

where \(\beta_{0,j}\) is the diameter-independent growth rate for group
\(j\); \(\text{dbh}_{ij}\) is the DBH the focal tree at the earlier
census; \(\beta_{\text{dbh},j}\) is the amount of the growth rate
changed depending on diameter for group \(j\); \(\text{BA}_{ijk}\) is
the sum of the basal area of all trees of competitor species group
\(k\); \(\lambda_{jk}\) is the change in growth for individuals of group
\(j\) from nearby competitors of group \(k\); and \(\epsilon_{ij}\) is a
random error term distributed \(\text{Normal}(0, \sigma^2)\). They
estimate all parameters via Bayesian linear regression while exploiting
Normal/Inverse Gamma conjugacy to derive closed-form solutions to all
posterior distributions\footnote{See S1 Appendix of
  \citet{allen_permutation_2020}, available at
  \url{https://doi.org/10.1371/journal.pone.0229930.s004}}. These
closed-form solutions for the posterior distributions are in contrast to
approximations of all posteriors via computationally expensive Markov
Chain Monte Carlo algorithms.

@@

We introduce the \texttt{forestecology} R package providing methods and
data for forest ecology model fitting and assessment, available on CRAN
(\url{https://cran.r-project.org/web/packages/forestecology/index.html})
and on GitHub (\url{https://github.com/rudeboybert/forestecology}). The
package implements all aspects of the model in Equation \ref{eq:model}:
model fitting and generating fitted/predicted values, evaluating the
effect of competitor species identity using permutation tests, and
evaluating model performance using spatial cross-validation.

@@

Currently the package only implements the Bayesian linear regression
model of tree growth based on neighborhood competition detailed in
Equation \ref{eq:model}. As we demonstrate in Section
\ref{model-fit-predict} however, the fitting of this model is
self-contained in a single function \texttt{comp\_bayes\_lm()}. This
function returns an object of S3 class type \texttt{comp\_bayes\_lm}
with generic methods implemented to print, make predictions using, and
plot all results. Therefore the package can be modularly extended to fit
other models as long as they are coded into a function similar to
\texttt{comp\_bayes\_lm()} and has equivalent generic methods
implemented.

@@

We demonstrate the \texttt{forestecology} package's functionality on
data from the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot, located at the Smithsonian's National Zoo and
Conservation Biology Institute in Front Royal, VA, USA
\citep{bourg_initial_2013}. The 25.6 ha (640 x 400 m) plot is located at
the intersection of three of the major physiographic provinces of the
eastern US---the Blue Ridge, Ridge and Valley, and Piedmont
provinces---and is adjacent to the northern end of Shenandoah National
Park.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(lubridate)}
\KeywordTok{library}\NormalTok{(sf)}
\KeywordTok{library}\NormalTok{(patchwork)}
\KeywordTok{library}\NormalTok{(forestecology)}
\KeywordTok{library}\NormalTok{(blockCV)}

\CommentTok{# Resolve conflicting functions}
\NormalTok{filter <-}\StringTok{ }\NormalTok{dplyr}\OperatorTok{::}\NormalTok{filter}
\NormalTok{select <-}\StringTok{ }\NormalTok{dplyr}\OperatorTok{::}\NormalTok{select}
\end{Highlighting}
\end{Shaded}

\hypertarget{compute-growth}{%
\subsection{Step 1: Compute the growth of trees based on census
data}\label{compute-growth}}

The first step is to compute the growth of trees using data from two
censuses. \texttt{compute\_growth()} computes average annual growth
assuming census data that roughly follows ForestGEO standards. Despite
such standards, minor variations will still exist between sites, thereby
necessitating some data wrangling and checking. For example, the SCBI
site records all diameters at breast height (DBH) in millimeters
\citep{bourg_initial_2013}, whereas the Michigan Big Woods site records
them in centimeters \citep{allen_michigan_2020}.

We first load both 2008 and 2014 SCBI census data \texttt{.csv} files as
they existed on GitHub on November 20, 2020
\citep{gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020} and perform
some data wrangling to both data sets. We then only consider a 9 ha
subsection of the 25.6 ha of the SCBI site, \texttt{gx} from 0--300
instead of 0--400 and \texttt{gy} from 300--600 instead of 0--640, in
order to speed up computation for purposes of this example.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{census_}\DecValTok{2013}\NormalTok{_scbi <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"scbi.stem2.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(stemID, sp, }\DataTypeTok{date =}\NormalTok{ ExactDate, gx, gy, dbh, codes, status) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \CommentTok{# Convert date from character to date}
    \DataTypeTok{date =} \KeywordTok{mdy}\NormalTok{(date),}
    \CommentTok{# Convert dbh to be in cm}
    \DataTypeTok{dbh =} \KeywordTok{as.numeric}\NormalTok{(dbh)}\OperatorTok{/}\DecValTok{10}
\NormalTok{  ) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(gx }\OperatorTok{<}\StringTok{ }\DecValTok{300}\NormalTok{, }\KeywordTok{between}\NormalTok{(gy, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{))}

@@

These two data frames are then supplied as arguments to
\texttt{compute\_growth()}, along with the \texttt{id} argument that
specifies the variable that uniquely identifies each tree-stem. Note
furthermore that we discard all resprouts in the later census (those
with \texttt{code\ ==\ R}), since we are only interested in the diameter
growth of surviving, and not resprouted, stems.

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes variables \texttt{growth}, the average annual
growth in DBH (cm \(\cdot\) y\textsuperscript{-1}) for all stems that
were alive at both time points, and \texttt{geometry}, the \texttt{sf}
package's encoding of geolocations of type
\texttt{\textless{}POINT\textgreater{}}. In addition the species
variable \texttt{sp} is returned as a factor.\footnote{In our spatial
  cross-validation algorithm in Section \ref{spatial-cross-validation}
  issues can occur when rare species do not occur in the training set,
  but then are encountered in the test set. This risk is mitigated by
  representing \texttt{sp} as a factor variable, which has a complete
  list of all levels of the categorical variable.}

@@

The next step is to add additional spatial information to
\texttt{growth\_scbi}. The first element we add is a ``buffer region''
to the periphery of the study region. Since some of our model's
explanatory variables are cumulative, we must ensure that all trees
being modeled are not biased to have different neighbor structures. This
is of concern for trees at the boundary of study regions, for which all
neighbors will not be included in the censused stems. In order to
account for such edge effects, only trees that are not part of this
buffer region, i.e.~are part of the interior of the study region, will
have their growth modeled \citep{waller_applied_2004}.

Our model of interspecific competition relies on a spatial definition of
who the competitor trees are for focal trees of interest: all trees
within a distance \texttt{comp\_dist} of a focal tree are considered its
competitors. In our case we set this value at 7.5m, a value informed by
other studies \citep[\citet{uriarte_spatially_2004},
\citet{canham_neighborhood_2006}]{canham_neighborhood_2004}. Using this
value along with a manually constructed \texttt{sf} object
representation of the study region's boundary, we apply the
\texttt{add\_buffer\_variable()} to \texttt{growth\_scbi} to add a
\texttt{buffer} boolean variable. All trees with \texttt{buffer} as
\texttt{FALSE} will be our focal trees whose growth will be modeled,
whereas those with \texttt{TRUE} will only be considered as competitor
trees.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Define competitive distance range}
\NormalTok{comp_dist <-}\StringTok{ }\FloatTok{7.5}

\CommentTok{# Manually construct study region boundary}
\NormalTok{study_region_scbi <-}\StringTok{ }\KeywordTok{tibble}\NormalTok{(}
  \DataTypeTok{x =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{300}\NormalTok{, }\DecValTok{300}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{),}
  \DataTypeTok{y =} \KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{, }\DecValTok{600}\NormalTok{, }\DecValTok{300}\NormalTok{)}
\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{sf_polygon}\NormalTok{()}

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate
out-of-sample model error. Conventional cross-validation algorithms
assign observations to folds by randomly resampling individual
observations. However, many of these algorithms assume that the
observations are independent. In the case of forest census data,
observations exhibit spatial autocorrelation. We therefore incorporate
this spatial dependence into the cross-validation algorithm with our
spatial blocks of trees \citep[
\citet{pohjankukka_estimating_2017}]{roberts_cross-validation_2017}.

In the example below, we first manually define four folds that partition
the study region as an \texttt{sf} object. We then use the output of the
\texttt{spatialBlock()} function from the \texttt{blockCV} package to
associate each tree in \texttt{growth\_scbi} to the correct fold
\texttt{foldID} \citep{valavi_blockcv_2019}. \footnote{In the Supporting
  Information we present an example where the folds themselves are also
  created automatically using \texttt{spatialBlock()} given a specified
  \texttt{cv\_block\_size}, as opposed to manually as in the example.}

@@

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of
adding these two elements of spatial information to
\texttt{growth\_scbi}. The location of each tree is marked with an
integer indicating its fold, where the folds are marked with solid
lines. The color of each digit indicates whether the tree is part of the
buffer region (and thus will only be considered as a competitor tree in
our model) or is part of the interior of the study region (and thus is a
focal tree whose growth is of modeled interest).

@@

The next step is to identify all focal trees and their corresponding
competitor trees. More specifically, identify all trees that are not
part of the buffer region, have a valid \texttt{growth} measurement, and
have at least one neighbor within 7.5m.
\texttt{create\_focal\_vs\_comp()} returns a new data frame of type
\texttt{sf}. On top of previously detailed arguments \texttt{comp\_dist}
and \texttt{id}, \texttt{create\_focal\_vs\_comp()} also requires an
\texttt{sf} object representation of the spatial cross-validation
blocks/folds as seen in Section \ref{spatial-information}. We present
the resulting data frame below with the \texttt{foldID} variable omitted
for compactness.

@@

The resulting \texttt{focal\_vs\_comp\_scbi} has 6296 rows, representing
the subset of the 7954 trees in \texttt{growth\_scbi} that will be
considered as focal trees. Two new variables \texttt{focal\_ID} and
\texttt{focal\_sp} relate to tree-stem identification and species
information. Most notably however is a new variable \texttt{comp} which
contains information on all competitor trees saved in \texttt{tidyr}
package list-column format \citep{tidyr_package}. We flatten the
\texttt{comp} list-column for the tree with \texttt{focal\_ID} 4 in the
first row, here a \texttt{tibble\ {[}20\ ×\ 4{]}}, into regular columns
using \texttt{unnest()} from the \texttt{tidyr} package.

We observe that for this focal tree, we have 4 variables of information
on its 20 competitor trees: their unique tree-stem ID number, their
distance to the focal tree (all \(\leq\) 7.5), their species, and their
basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) where \(DBH\) is the value
from the earlier of the two censuses in cm. Saving our focal versus
competitor information in list-column minimizes redundancy since we do
not repeat information on the focal tree 20 times. The spatial
distribution of these trees is visualized in Figure
\ref{fig:scbi-focal-vs-comp-map}: the dashed circle extends 7.5 m away
from the focal tree while all 20 competitor trees are within this
circle.

@@

@@

The final step is to fit a model for the growth of all focal trees. We
fit the competition Bayesian linear regression model outlined in
Equation \ref{eq:model} using \texttt{comp\_bayes\_lm()}, which has an
option to specify prior distributions on all parameters of interest
(chosen to be the defaults specified in \texttt{?comp\_bayes\_lm}).

@@

The returned \texttt{comp\_bayes\_lm\_scbi} output is an object of S3
class type \texttt{comp\_bayes\_lm} containing the posterior values of
all parameters in our competition Bayesian linear regression. This class
of object includes three generic methods. First, the generic for
\texttt{print()} displays the names of all prior \& posterior parameters
along with the model formula:
